---
title: "Gun Violence Data Preparation"
author: "Chinedu, Charles, Coffy"
date: "11/27/2021"
output: html_document
---


```{r load-libraries, message=FALSE}
# Load Libraries
library(tidyverse)
library(lubridate)
library(stringr)
library(leaflet)
library(sp)
library(ggmap)
library(leaflet.extras)
library(htmltools)
```


<center> <h2> DATA PREPARATION </h2> </center>

**Data Collection**: We downloaded eight(8) years of data from [GVA](https://www.gunviolencearchive.org/) and combined all into a single dataframe. The data collected does not have longitute and latitude for the locations, so we used the google API to obtain the lon and lat values for all 3335 observations.
```{r read-file, message=FALSE}
# read the data
mass_shooting2014 <- read_csv("https://raw.githubusercontent.com/chinedu2301/data607-project-gun-violence/main/Data/mass%20shootings%202014.csv")
mass_shooting2015 <- read_csv("https://raw.githubusercontent.com/chinedu2301/data607-project-gun-violence/main/Data/mass%20shootings%202015.csv")
mass_shooting2016 <- read_csv("https://raw.githubusercontent.com/chinedu2301/data607-project-gun-violence/main/Data/mass%20shootings%202016.csv")
mass_shooting2017 <- read_csv("https://raw.githubusercontent.com/chinedu2301/data607-project-gun-violence/main/Data/mass%20shootings%202017.csv")
mass_shooting2018 <- read_csv("https://raw.githubusercontent.com/chinedu2301/data607-project-gun-violence/main/Data/mass%20shootings%202018.csv")
mass_shooting2019 <- read_csv("https://raw.githubusercontent.com/chinedu2301/data607-project-gun-violence/main/Data/mass%20shootings%202019.csv")
mass_shooting2020 <- read_csv("https://raw.githubusercontent.com/chinedu2301/data607-project-gun-violence/main/Data/mass%20shootings%202020.csv")
mass_shooting2021 <- read_csv("https://raw.githubusercontent.com/chinedu2301/data607-project-gun-violence/main/Data/mass%20shootings%202021.csv")
```


```{r}
# Combine the data-sets from 2014 to 2021
mass_shooting <- rbind(mass_shooting2014, mass_shooting2015, mass_shooting2016, mass_shooting2017,mass_shooting2018, mass_shooting2019, mass_shooting2020, mass_shooting2021)
head(mass_shooting, n=10)
```
```{r}
# rename the columns
headers <- c("Incident_ID", "Incident_date", "State", "City", "Address", "No_killed", "No_injured",
             "Operations")
colnames(mass_shooting) <- headers

# Combine the Address, City, and State columns using str_c
mass_shooting$Geo_Address <- str_c(mass_shooting$Address, ", ", mass_shooting$City, ", ", mass_shooting$State)

# Convert the Incident_date column from character to date format
mass_shooting$Incident_date <- mdy(mass_shooting$Incident_date)

# Select columns needed for analysis
mass_shooting <- mass_shooting %>% select(Incident_ID, Incident_date, State, City, Address,
                                          No_killed, No_injured, Geo_Address, )

# Check the head of the data set
head(mass_shooting, n=10)
```
```{r}
# Glimpse
glimpse(mass_shooting)
```

```{r}
# use google API to get the Lon and Lat for each of the Geo_Adresses
mass_shooting_df <- mass_shooting %>% as.data.frame()
mass_shooting2 <- mass_shooting_df %>% mutate_geocode(Geo_Address)
head(mass_shooting2)
```
The google Geocode API timed-out and failed to return lon and lat values for 4 locations
```{r}
# convert the data frame back to tibble
mass_shooting2_tib <- as_tibble(mass_shooting2)

# filter for observations with NA values for lon
mass_shooting_lonlat_na <- mass_shooting2_tib %>% filter(is.na(lon))
mass_shooting_lonlat_na
```

```{r}
# Obtain lon and lat for these four locations
mass_shooting_lonlat_missing <- mass_shooting_lonlat_na %>% select(-c("lon", "lat")) %>%
  mutate_geocode(Geo_Address)
mass_shooting_lonlat_missing
```

```{r}
# replace the NA value with the lat and lon for that location
mass_shooting_aliceville <- geocode("1st Ave and 15th St, Aliceville, Alabama")
mass_shooting_lonlat_missing <- mass_shooting_lonlat_missing %>% replace_na(mass_shooting_aliceville)
mass_shooting_lonlat_missing
```

```{r}
# Replace the observations with NA values in mass_shooting2 with their corresponding values
for(i in mass_shooting_lonlat_missing$Incident_ID){
  mass_shooting2[mass_shooting2$Incident_ID == i, ] = mass_shooting_lonlat_missing[mass_shooting_lonlat_missing$Incident_ID == i, ]
}
```

```{r}
# Check for any observations with NA values for the lon and lat
mass_shooting2 %>% filter(is.na(lat) | is.na(lon))
```

No observation have NA values for the lon or lat.

```{r}
# Check the head of the dataframe
head(mass_shooting2, n = 10)
```
```{r}
# Check the glimpse for the final dataset
glimpse(mass_shooting2)
```



```{r}
write.csv(mass_shooting2, file = "mass_shooting.csv")
```



